Perturbative probability distribution of the average Ising magnetization in 4 dimensions.

### 2020-04-18 paper/paper.tex
Proofread 1

\title{One-loop PDF of the average Ising magnetization in 4 dimensions.}
\author[a]{Andrea Allais}
\affiliation[a]{Independent scholar}
\emailAdd{allais.andrea@gmail.com}
\abstract{We compute at one loop in perturbation theory the probability density
function of the average magnetization $\bar{\phi}$ of the Ising model on both
the 4-torus and the 4-sphere. The perturbative expansion covers the entire
symmetric phase and a neighborhood of the critical point. We find that, at the
critical point, the PDF has the asymptotic behavior $p(\bar{\phi})\sim
\exp(-f(R) \bar{\phi}^4)$, and hence the critical value of the Binder cumulant
is $U = 1 - \frac{4\,\Gamma(5/4)^2}{ 3\,\Gamma(3/4)^2}$. We validate our
results by comparison with Monte Carlo simulation.}
\begin{document}

@@

The average magnetization $\bar{\phi} = \frac{1}{V} \sum_{i} \sigma_i$ of the
Ising model is the simplest observable in Monte Carlo simulations. For large
system size, the probability distribution of $\bar{\phi}$ changes qualitatively
between the two phases of the model. In the symmetric phase, if the correlation
length $\xi$ is large in lattice units, but small compared to the system size
$L$, the distribution is well approximated by a normal distribution centered at
0. In the broken symmetry phase, still for $1 \ll \xi \ll L$, the distribution
is well approximated by a bimodal superposition of two normals centered at
non-zero values $\pm \bar{\phi}_0$.

Near the critical point $1 \ll L \ll \xi$ the probability distribution of
$\bar{\phi}$ is a non-trivial function of $\bar{\phi}$, at least in two and
three dimensions.  It is not immediately clear if the same is true in four
dimensions as well.  Since the $\phi^4$ field theory that describes the Ising
critical point in four dimensions becomes weakly coupled in the infrared, one
may expect the probability distribution to be a single zero-centered normal at
the critical point as well.

To clarify this issue, we compute the log pdf of $\bar{\phi}$ at one-loop in
$4$ dimensions, and show that for very large system size, the critical
probability distribution is not normal, but rather of the form
$-\log p(\bar{\phi})\sim f(L) \bar{\phi}^4$. We find that corrections to this
form vanish very slowly with increasing system size, like $1/\log L$.

@@

Thus we consider the $\phi^4$ field theory with action:
\begin{equation}
  S[\phi] \equiv \int \dd^4 x\ \left(
    \frac{1}{2} \phi\left(\Delta + r_0\right)\phi +
    \frac{1}{24} u_0 \phi^4\right)
\end{equation}
defined on a 4-torus of radius $R$, i.e. $x_i\sim x_i + 2\pi R$. In this field
theory context, the pdf of the average magnetization is defined as:
\begin{equation}
    \label{eq:pdf_definition}
    p(\bar\phi) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
    \delta\left(\bar{\phi} - \frac{1}{V}\int\dd^4 x\ \phi\right)
    e^{-S[\phi]}\,,
\end{equation}
and at one loop it evaluates to:
\begin{equation}
\label{eq:log_pdf}
\begin{split}
    -\log p(\bar{\phi}) = \mathrm{const.} +  V\Bigg(
  &\frac{1}{2} \frac{\bar{\phi}^2}{R^2} \left(
    r R^2 + \frac{g}{6} \left(f_1\left(r R^2\right) - 1\right)
    + O\left(g^2\right)\right) + \\
  & \frac{2\pi^2}{9}g\bar{\phi}^4\left(
    1 - \frac{g}{2}f_2\left(r R^2\right)
    + O\left(g^2\right)\right) + \\
  & \frac{8\pi^4}{81} g^3 \bar{\phi}^6 R^2\left(
    f_3\left(r R^2\right)
    + O\left(g\right)\right) + O\left(g^4\bar{\phi}^8\right) \Bigg)\,,
\end{split}
\end{equation}
where the coefficients $r$ and $g$ are the renormalized counterparts to $r_0$
and $u_0$. An explicit expression for the functions $f_1$, $f_2$ and $f_3$ is
reported in the appendix (\ref{eq:f_functions}), and they are plotted in
fig.~\ref{fig:f_functions}.

@@

Eq. (\ref{eq:log_pdf}) is our central result, and we now describe it
at lenght. We first illustrate its domain of validity and its dependence on the
renormalized couplings $r$, $g$ on a torus of fixed radius $R$, then describe
its dependence on $R$ at fixed bare couplings $r_0$, $u_0$.

The dimension-2 coupling $r$ controls the cross-over between the symmetric and
broken-symmetry phases. The expression (\ref{eq:log_pdf}) is valid for
all positive values of $r$, and also for negative values, provided that $rR^2
\gtrsim -1/2$.  For lower values of $r$, additional modes of the laplacian
condense beside the zero-mode, and this invalidates the perturbative expansion.
. This becomes manifest in a divergence of the functions $f_i$ as their
argument approaches -1.  On the other hand, $f_1$, $f_2$, $f_3$ all go to zero
as their argument approaches positive infinity, and hence for large $r$ the
probability distribution of $\bar{\phi}$ is a zero-centered normal:
\begin{equation}
    p(\bar{\phi}) \sim N e^{- \frac{1}{2} V r \bar{\phi}^2}
  \quad \text{for} \quad
  rR^2 \gg 1\,.
\end{equation}

@@

These can be integrated and combined with the condition $\mu^2 = r + R^{-2}$ to
obtain a system of equations\footnote{There are of course many alternative,
arguably simpler, solutions that differ by subleading orders in an expansion in
$g(R_1)$. The one displayed here is the exact solution to
(\ref{eq:renormalization_group}).} connecting the renormalized couplings at two
different values of $R$:
\begin{equation}
  \frac{g(R_1)}{g(R_2)} = \frac{r(R_1)^3}{r(R_2)^3} = 
  1 - \frac{1}{2} g(R_1)\log\frac{r(R_2) + R_2^{-2}}{r(R_1) + R_1^{-2}}\,.
\end{equation}

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{renormalization_group.png}
\end{center}
    \caption{\label{fig:renormalization_group} A few solutions to
    (\ref{eq:renormalization_group}) with $g(R_1) = 0.45$.  In the left plot,
    the shaded region shows where the perturbative expansion becomes invalid.
    Similarly in the right plot, the lines become dotted oustside of the
    perturbative region. Note how the sign of $r$ is preserved, $r = 0$ is a
    solution, and $g$ goes to $0$ for $R_2 \gg R_1$.}
\end{figure}

@@

Thus we conclude that, for sufficiently large system size: 
\begin{equation}
    \label{eq:log_pdf_critical_point}
    -\log p(\phi) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,,
\end{equation}
with subleading terms vanishing like $1 / \log(R)$.

In Monte Carlo simulations, the qualitative behavior of the distribution of the
magnetization is often characterized by measuring the so-called Binder
cumulant \cite{Binder1981}:
\begin{equation}
    U = 1 - \frac{\left\langle \bar \phi^4 \right \rangle}
    {3\left\langle \bar \phi^2 \right \rangle^2},
\end{equation}
which is constructed to be independent of the overall scale of $\bar{\phi}$,
and to be zero if $\bar{\phi}$ is normally distributed. From
(\ref{eq:log_pdf_critical_point}) we conclude that, on a 4-Torus, at the
critical point\footnote{ Incidentally, we note that this result is not in
contraddiction with the claim in the original paper \cite{Binder1981} that the
critical Binder cumulant in 4 dimensions is zero, because in the original paper
the cumulant is not computed from the total magnetization, but rather from the
average magnetization of blocks of spins much smaller than the overall system
size.}:
\begin{equation}
    \label{eq:critical_binder_cumulant}
    U = 1 - \frac{4\,\Gamma\left(\frac{5}{4}\right)^2}
    {3\,\Gamma\left(\frac{3}{4}\right)^2} 
    = 0.27052\ldots
\end{equation}

@@

\section{Derivation on the 4-torus}

@@

First of all we observe that the average magnetization
$\bar{\phi}$ is proportional to the zero momentum mode of the field $\phi$:
\begin{equation}
    \phi(x) = \frac{1}{V} \sum_{n \in \mathbb{Z}^4} \phi_{n} 
    e^{i \frac{n\cdot x}{R}}\,;\quad \bar{\phi} = \frac{\phi_{0}}{V}\,.
\end{equation}

Thus, because of the delta function in (\ref{eq:pdf_definition}), the component
$\phi_{0}$ becomes an external field, whereas the other components $\phi_{n}$
with are still part of the functional integral. Separating the zero momentum
mode from the other modes in the action yields:
\begin{equation}
\begin{split}
    S(\bar{\phi}, \phi) =& 
    V \left(\frac{1}{2} r_0 \bar{\phi}^2 + \frac{u_0}{24} \bar{\phi}^4\right) +
    \frac{u_0}{4V} \bar{\phi}^2 \sum_{n} \phi_{n} \phi_{-n} + \\
    &\frac{u_0}{6V^2} \bar{\phi} 
        \sum_{n_1, n_2} \phi_{n_1} \phi_{n_2} \phi_{-n_1 -n_2} + \\
    &\frac{u_0}{24V^3} \sum_{n_1, n_2, n_3} \phi_{n_1} \phi_{n_2} 
        \phi_{n_3} \phi_{-n_1 -n_2 -n_3}\,,
\end{split}
\end{equation}
where all summations now exclude $n = (0, 0, 0, 0)$.

We also introduce renormalized couplings
\begin{equation}
    u_0 = u(1 + u\delta u)\,;\quad r_0 = r(1 + u \delta r)\,,
\end{equation}
and we obtain the following edges and vertices in the diagrammatic expansion:
\begin{displaymath}
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[densely dashed, thick] (-1, -1) -- (0, 0) -- (-1, 1);
    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
    \node[above right] at (1, 1) {$n_1$};
    \node[below right] at (1, -1) {$n_2$};
\end{tikzpicture}
    -\frac{u\bar{\phi}^2}{4V}\delta_{n_1 + n_2}
    \quad\quad
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[densely dashed, thick] (-1, -1) -- (0, 0);
    \draw[thick] (0, 0) -- (-1, 1);
    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
    \node[above left] at (-1, 1) {$n_1$};
    \node[above right] at (1, 1) {$n_2$};
    \node[below right] at (1, -1) {$n_3$};
\end{tikzpicture}
    -\frac{u\bar{\phi}}{6V^2}\delta_{n_1 + n_2 + n_3}
\end{displaymath}
\begin{displaymath}
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[thick] (-1, -1) -- (1, 1);
    \draw[thick] (-1, 1) -- ( 1, -1);
    \node[above left] at (-1, 1) {$n_1$};
    \node[above right] at (1, 1) {$n_2$};
    \node[below right] at (1, -1) {$n_3$};
    \node[below left] at (-1, -1) {$n_4$};
\end{tikzpicture}
    -\frac{u}{24V^3}\delta_{n_1 + n_2 + n_3 + n_4}
\quad\quad
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[thick] (-1, 0) -- (1, 0);
    \node[left] at (-1, 0) {$n_1$};
    \node[right] at (1, 0) {$n_2$};
\end{tikzpicture}
    \quad
    \frac{V}{n_1^2R^{-2} + r}\delta_{n_1 + n_2}
\end{displaymath}
plus additional vertices associated with the counterterms $\delta r$, $\delta
u$ which we do not list for brevity.

@@

Retaining only one-loop diagrams, and employing a heath kernel regulator we have:
\begin{equation}
\begin{split}
    -\log p(\bar{\phi}) = \mathrm{const} + V \Bigg( 
    &\frac{1}{2} \bar{\phi}^2\left( r + u \left(\delta r + \frac{1}{2} I_1\right) + O(u^2)\right) + \\
    &\frac{u}{24} \bar{\phi}^4 \left(1 + u \left(\delta u - \frac{3}{2} I_2\right) + O(u^2)\right) + \\
    &\frac{u^3}{48}\bar{\phi}^6\left(I_3 + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
\end{split}
\end{equation}
where
\begin{equation}
    I_k = \frac{1}{V} \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    \frac{e^{-s(r + n^2 R^{-2})}}{\left(r + n^2 R^{-2}\right)^k} = 
    \frac{1}{(k - 1)!}\left(-\frac{\partial}{\partial r} - s\right)^k I_1
\end{equation}

The summation $I_1$ can be simplified substantially using the following trick:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    e^{-t(r + n^2 R^{-2})}\\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\left(\sum_{m = -\infty}^{\infty} e^{-t m^2 R^{-2}}\right)^4 - 1\right) \\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\theta\!\left(t R^{-2}\right)^4 - 1\right)\,,
\end{split}
\end{equation}
where:
\begin{equation}
    \label{eq:theta}
    \theta(z) = \theta_3(0, e^{-z}) = \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
\end{equation}
Here $\theta_3$ is a Jacobi theta function, but we will not need any of
its special properties beyond its asymptotic behavior, which can be easily
obtained from the definition.

We now extract the divergent and finite parts of $I_1$ as $s\to 0$ by
subtracting under the integral a function that has the same asymptotic behavior
as the integrand for $t \to 0$, but whose integral can be computed in closed
form. In order to do that, we need the asymptotic behavior of $\theta(z)$ for
$z\to0$, which can be obtained from its definition using the Euler-Mclaurin
formula:
\begin{equation}
    \theta(z) \sim \sqrt{\frac{\pi}{z}} + o(z^k) \quad\mathrm{for}\, z\to0\,.
\end{equation}

@@

We set the counterterms to:
\begin{align}
    &\delta r = -\frac{1}{32\pi^2}\left(\frac{1}{s} + r\left(\log s \mu^2 + \gamma_E - 1\right)\right)\,,\\
    &\delta u = -\frac{3}{32\pi^2}\left(\log s\mu^2 +\gamma_E + 1\right)\,,
\end{align}
from which the renormalization group equations (\ref{eq:renormalization_group})
can be obtained, and we have:
\begin{equation}
\begin{split}
    -\log p(\bar{\phi}) =& \mathrm{const}+ V \Bigg( \\
    &\frac{1}{2} \bar{\phi}^2\left( r + \frac{u}{32\pi^2}  \left(\left(\frac{1}{R^2} f(rR^2, \mu R) + r - \mu^2\right)\right) + O(u^2)\right) + \\
    &\frac{u}{24} \bar{\phi}^4 \left(1 + \frac{3u}{32\pi^2} f^{(1, 0)}(rR^2, \mu R) + O(u^2)\right) + \\
    &\frac{u^3}{48}\bar{\phi}^6\left(\frac{R^2}{32\pi^2}f^{(2, 0)}(rR^2, \mu R) + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
\end{split}
\end{equation}
from which (\ref{eq:log_pdf}) is obtained by setting $g = \frac{3 u}{16
\pi^2}$ and $\mu^2 = r + R^{-2}$. This last choice is motivated as follows. The
subtraction in (\ref{eq:f_definition}) is similar to:
\begin{equation}
    \int_0^{\infty} \dd z\, \frac{e^{-a z} - e^{-b z}}{z} = \log \frac{b}{a}\,;
\end{equation}
when the asymptotic behavior of the two terms for large $z$ is not well
matched, the integral becomes large in magnitude, making the perturbative
expansion less reliable. With the choice $\mu^2 = r + R^{-2}$, both terms in
(\ref{eq:f_definition}) have the same asymptotic behavior $\sim e^{-z(\bar{r} +
1)}$, thus avoiding the large log problem over the widest possible range of
parameters.

@@

Curved manifolds allow for an additional renormalizable coupling: 
\begin{equation}
    S += \int \dd^4 x\ \frac{1}{2}\xi_0 \phi^2 \mathcal{R}\,,
\end{equation}
where $\mathcal{R}$ is the scalar curvature and $\xi_0$ is a dimensionless bare
coupling. On a static manifold, this coupling is equivalent to a mass term.
The free theory is Weyl invariant if $\xi_0 = \frac{1}{6}$. In the
presence of interactions, the coupling needs to be renormalized as well, and
its renormalized counterpart $\xi$ becomes a running coupling. At one loop, the
renormalization group equation for $\xi$ is:
\begin{equation}
    \mu \frac{\dd \xi}{\dd \mu} =  \frac{1}{3} g \left(\xi - \frac{1}{6}\right)\,.
\end{equation}

From this experssion it seems that $\xi$ can be set to the critical value
$\frac{1}{6}$ at all energy scales. However, this turns out to be an illusion:
at higher orders in perturbation theory the renormalization group equation
becomes inhomogeneous \cite{Brown1980}. Therefore, on a curved manifold, $\xi$
is simply another free parameter of the scalar field theory.

>>>>

\title{Perturbative probability distribution of the average Ising magnetization
in 4 dimensions.}

\author[a]{Andrea Allais}
\affiliation[a]{Independent scholar}
\emailAdd{allais.andrea@gmail.com}

\abstract{We compute, at one loop in perturbation theory, the probability density
function of the average magnetization $\bar{\phi}$ of the Ising model, on both
the 4-torus and the 4-sphere. The perturbative expansion covers the entire
symmetric phase and a neighborhood of the critical point. We find that, at the
critical point, the PDF has the asymptotic behavior $p(\bar{\phi})\sim
\exp(-c \bar{\phi}^4)$, and hence the critical value of the Binder cumulant
is $U = 1 - \frac{4\,\Gamma(5/4)^2}{ 3\,\Gamma(3/4)^2}$. We validate our
results by comparison with Monte Carlo simulation.}
\begin{document}

@@

The average magnetization $\bar{\phi} = \frac{1}{V} \sum_{i} \sigma_i$ of the
Ising model is the simplest observable in Monte Carlo simulations. When the
system size $L$ is large, the probability distribution of $\bar{\phi}$ changes
qualitatively between the two phases of the model. In the symmetric phase,
\textit{i. e.} when the correlation length $\xi$ is large in lattice units, but
small compared to the system size, the distribution is well approximated by a
zero-mean normal. In the broken symmetry phase, still for $1 \ll \xi \ll L$,
the distribution is well approximated by a mixture of two normals centered at
non-zero values $\pm \bar{\phi}_0$.

Near the critical point, \textit{ie.} $1 \ll L \ll \xi$, the probability
distribution of $\bar{\phi}$ is a non-trivial function of $\bar{\phi}$, at
least in two and three dimensions.  It is not immediately clear if the same is
also true in four dimensions. Since the field theory that describes the Ising
critical point in four dimensions becomes weakly coupled at low energy, one may
expect the probability distribution of $\bar{\phi}$ to remain a zero-centered
normal at the critical point as well as in the symmetric phase.

To clarify this issue, we compute the logarithm of the PDF of $\bar{\phi}$ at
one-loop in $4$ dimensions, and show that, for very large system size, the
critical probability distribution is not normal, but rather of the form $-\log
p(\bar{\phi})\sim f(L) \bar{\phi}^4$. We find that corrections to this form
vanish very slowly with increasing system size, like $1/\log L$.

@@

Thus we consider the $\phi^4$ field theory with action:
\begin{equation}
  S[\phi] \equiv \int \dd^4 x\ \left(
    \frac{1}{2} \phi(x)\left(\Delta + r_0\right)\phi(x) +
    \frac{1}{24} u_0 \phi(x)^4\right)
\end{equation}
defined on a 4-torus of radius $R$, i.e. $x_i\sim x_i + 2\pi R$. In this field
theory context, the PDF of the average magnetization is defined as:
\begin{equation}
    \label{eq:pdf_definition}
    p(\bar\phi) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
    \delta\left(\bar{\phi} - \frac{1}{V}\int\dd^4 x\ \phi(x)\right)
    e^{-S[\phi]}\,,
\end{equation}
and at one loop it evaluates to:
\begin{equation}
\label{eq:log_pdf}
\begin{split}
    -\log p(\bar{\phi}) = \mathrm{const.} +  V\Bigg(
  &\frac{1}{2} \frac{\bar{\phi}^2}{R^2} \left(
    r R^2 + \frac{g}{6} \left(f_1\left(r R^2\right) - 1\right)
    + O\left(g^2\right)\right) + \\
  & \frac{2\pi^2}{9}g\bar{\phi}^4\left(
    1 - \frac{g}{2}f_2\left(r R^2\right)
    + O\left(g^2\right)\right) + \\
  & \frac{8\pi^4}{81} g^3 \bar{\phi}^6 R^2\left(
    f_3\left(r R^2\right)
    + O\left(g\right)\right) + O\left(g^4\bar{\phi}^8\right) \Bigg)\,,
\end{split}
\end{equation}
where the coefficients $r$ and $g$ are the renormalized counterparts to $r_0$
and $u_0$. An explicit expression for the functions $f_1$, $f_2$ and $f_3$ is
reported in section \ref{sect:derivation} (\ref{eq:f_functions}), and they are
plotted in
fig.~\ref{fig:f_functions}.

@@

Eq. (\ref{eq:log_pdf}) is our central result, and we now describe it
at length. We first illustrate its domain of validity and its dependence on the
renormalized couplings $r$, $g$ on a torus of fixed radius $R$, then describe
its dependence on $R$ at fixed bare couplings $r_0$, $u_0$.

The dimension-2 coupling $r$ controls the cross-over between the symmetric and
broken-symmetry phases. The expression (\ref{eq:log_pdf}) is valid for all
positive values of $r$, and also for negative values, provided that $rR^2
\gtrsim -1/2$.  For lower values of $r$, additional modes of the Laplacian
condense beside the zero-mode, and this invalidates the perturbative expansion.
This becomes manifest in a divergence of the functions $f_i$ as their argument
approaches -1.  On the other hand, $f_1$, $f_2$, $f_3$ all go to zero as their
argument approaches positive infinity, and hence for large $r$ the probability
distribution of $\bar{\phi}$ is a zero-centered normal:
\begin{equation}
    p(\bar{\phi}) \sim N e^{- \frac{1}{2} V r \bar{\phi}^2}
  \quad \text{for} \quad
  rR^2 \gg 1\,.
\end{equation}

@@

These can be integrated and combined with the condition $\mu^2 = r + R^{-2}$ to
obtain a system of equations\footnote{There are of course many alternative,
arguably simpler, solutions that differ by sub-leading orders in an expansion in
$g(R_1)$. The one displayed here is the exact solution to
(\ref{eq:renormalization_group}).} connecting the renormalized couplings at two
different values of $R$:
\begin{equation}
  \frac{g(R_1)}{g(R_2)} = \frac{r(R_1)^3}{r(R_2)^3} = 
  1 - \frac{1}{2} g(R_1)\log\frac{r(R_2) + R_2^{-2}}{r(R_1) + R_1^{-2}}\,.
\end{equation}

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{renormalization_group.png}
\end{center}
    \caption{\label{fig:renormalization_group} A few solutions to
    (\ref{eq:renormalization_group}) with $g(R_1) = 0.45$.  In the left plot,
    the shaded region shows where the perturbative expansion becomes invalid.
    Similarly in the right plot, the lines become dotted outside of the
    perturbative region. Note how the sign of $r$ is preserved, $r = 0$ is a
    solution, and $g$ vanishes for $R_2 \gg R_1$.}
\end{figure}

@@

Thus we conclude that, at the critical point, for sufficiently large system
size:
\begin{equation}
    \label{eq:log_pdf_critical_point}
    -\log p(\bar{\phi}) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,,
\end{equation}
with sub-leading terms vanishing like $1 / \log(R)$.

In Monte Carlo simulations, the qualitative behavior of the distribution of the
magnetization is often characterized by measuring the so-called Binder
cumulant \cite{Binder1981}:
\begin{equation}
    U = 1 - \frac{\left\langle \bar \phi^4 \right \rangle}
    {3\left\langle \bar \phi^2 \right \rangle^2},
\end{equation}
which is constructed to be independent of the overall scale of $\bar{\phi}$,
and to be zero if $\bar{\phi}$ is normally distributed. From
(\ref{eq:log_pdf_critical_point}) we conclude that, on a 4-Torus, at the
critical point\footnote{ Incidentally, we note that this result is not in
contradiction with the claim in the original paper \cite{Binder1981} that the
critical Binder cumulant in 4 dimensions is zero, because in the original paper
the cumulant is not computed from the total magnetization, but rather from the
average magnetization of blocks of spins much smaller than the overall system
size.}:
\begin{equation}
    \label{eq:critical_binder_cumulant}
    U = 1 - \frac{4\,\Gamma\left(\frac{5}{4}\right)^2}
    {3\,\Gamma\left(\frac{3}{4}\right)^2} 
    = 0.27052\ldots
\end{equation}

@@

\section{Derivation on the 4-torus}
\label{sect:derivation}

@@

First of all we observe that the average magnetization
$\bar{\phi}$ is proportional to the zero momentum mode of the field $\phi$:
\begin{equation}
    \phi(x) = \frac{1}{V} \sum_{n \in \mathbb{Z}^4} \phi_{n} 
    e^{i \frac{n\cdot x}{R}}\,;\quad \bar{\phi} = \frac{1}{V}\int\dd^4 x\ \phi(x) = \frac{1}{V}\phi_{n = 0}\,.
\end{equation}

Thus, because of the delta function in (\ref{eq:pdf_definition}), the component
$\phi_{0}$ becomes an external field, whereas the other components $\phi_{n}$
are still part of the functional integral. Separating the zero momentum mode
from the other modes in the action yields:
\begin{equation}
\begin{split}
    S(\bar{\phi}, \phi) =& 
    V \left(\frac{1}{2} r_0 \bar{\phi}^2 + \frac{u_0}{24} \bar{\phi}^4\right) +
    \frac{1}{2V} \sum_{n} \left(\frac{n^2}{R^2} + r_0 +
      \frac{1}{2} u_0\bar{\phi}^2 \right)\phi_{n} \phi_{-n} +\\
    &\frac{u_0}{6V^2} \bar{\phi} 
        \sum_{n_1, n_2} \phi_{n_1} \phi_{n_2} \phi_{-n_1 -n_2} +
    \frac{u_0}{24V^3} \sum_{n_1, n_2, n_3} \phi_{n_1} \phi_{n_2}
        \phi_{n_3} \phi_{-n_1 -n_2 -n_3}\,,
\end{split}
\end{equation}
where all summations now are over $\mathbb{Z}^4 \setminus \{0\}$.

We also introduce renormalized couplings
\begin{equation}
    u_0 = u(1 + u\delta u)\,;\quad r_0 = r(1 + u \delta r)\,,
\end{equation}
and we obtain the following edges and vertices in the diagrammatic expansion:
\begin{displaymath}
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[thick] (-1, 0) -- (1, 0);
    \node[left] at (-1, 0) {$n_1$};
    \node[right] at (1, 0) {$n_2$};
\end{tikzpicture}
    \quad
    \frac{V}{n_1^2R^{-2} + r}\delta_{n_1 + n_2}
\quad\quad
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[densely dashed, thick] (-1, -1) -- (0, 0) -- (-1, 1);
    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
    \node[above right] at (1, 1) {$n_1$};
    \node[below right] at (1, -1) {$n_2$};
\end{tikzpicture}
    -\frac{u\bar{\phi}^2}{4V}\delta_{n_1 + n_2}
\end{displaymath}
\begin{displaymath}
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[densely dashed, thick] (-1, -1) -- (0, 0);
    \draw[thick] (0, 0) -- (-1, 1);
    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
    \node[above left] at (-1, 1) {$n_1$};
    \node[above right] at (1, 1) {$n_2$};
    \node[below right] at (1, -1) {$n_3$};
\end{tikzpicture}
    -\frac{u\bar{\phi}}{6V^2}\delta_{n_1 + n_2 + n_3}
\quad\quad
\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
    \draw[thick] (-1, -1) -- (1, 1);
    \draw[thick] (-1, 1) -- ( 1, -1);
    \node[above left] at (-1, 1) {$n_1$};
    \node[above right] at (1, 1) {$n_2$};
    \node[below right] at (1, -1) {$n_3$};
    \node[below left] at (-1, -1) {$n_4$};
\end{tikzpicture}
    -\frac{u}{24V^3}\delta_{n_1 + n_2 + n_3 + n_4}
\end{displaymath}
plus additional vertices associated with the counterterms $\delta r$, $\delta
u$ which we do not list for brevity.

@@

Retaining only one-loop diagrams, and employing a heath kernel regulator we have:
\begin{equation}
\begin{split}
    -\log p(\bar{\phi}) = \mathrm{const} + V \Bigg( 
    &\frac{1}{2} \bar{\phi}^2\left( r + u \left(\delta r + \frac{1}{2} I_1\right) + O(u^2)\right) + \\
    &\frac{u}{24} \bar{\phi}^4 \left(1 + u \left(\delta u - \frac{3}{2} I_2\right) + O(u^2)\right) + \\
    &\frac{u^3}{48}\bar{\phi}^6\left(I_3 + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
\end{split}
\end{equation}
where
\begin{equation}
    I_k = \frac{1}{V} \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    \frac{e^{-s(n^2 R^{-2} + r)}}{\left(n^2 R^{-2} + r\right)^k} = 
    \frac{1}{(k - 1)!}\left(-\frac{\partial}{\partial r} - s\right)^{k-1} I_1
\end{equation}

The summation $I_1$ can be simplified substantially using the following trick:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    e^{-t(n^2 R^{-2} + r)}\\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\left(\sum_{m = -\infty}^{\infty} e^{-t m^2 R^{-2}}\right)^4 - 1\right) \\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\theta\!\left(t R^{-2}\right)^4 - 1\right)\,,
\end{split}
\end{equation}
where:
\begin{equation}
    \label{eq:theta}
    \theta(z) = \theta_3(0; e^{-z}) = \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
\end{equation}
Here $\theta_3$ is a Jacobi theta function, but we will not need any of
its special properties beyond its asymptotic behavior, which can be easily
obtained from the definition.

We now extract the divergent and finite parts of $I_1$ as $s\to 0$ by
subtracting under the integral a function that has the same asymptotic behavior
as the integrand for $t \to 0$, but whose integral can be computed in closed
form. In order to do that, we need the asymptotic behavior of $\theta(z)$ for
$z\to0$, which can be obtained from its definition using the Euler-Maclaurin
formula:
\begin{equation}
    \theta(z) \sim \sqrt{\frac{\pi}{z}} + o(z^k) \quad\mathrm{for}\, z\to0\,.
\end{equation}

@@

We set the counterterms to:
\begin{align}
    &\delta r = -\frac{1}{32\pi^2}\left(\frac{1}{s} + r\left(\log s \mu^2 + \gamma_E - 1\right)\right)\,,\\
    &\delta u = -\frac{3}{32\pi^2}\left(\log s\mu^2 +\gamma_E + 1\right)\,,
\end{align}
from which the renormalization group equations (\ref{eq:renormalization_group})
can be obtained, and we have:
\begin{equation}
\begin{split}
    -\log p(\bar{\phi}) =& \mathrm{const}+ V \Bigg(
    \frac{1}{2} \bar{\phi}^2\left( r + \frac{u}{32\pi^2R^2}  \left(f(rR^2, \mu R) + (r - \mu^2)R^2\right) + O(u^2)\right) + \\
    &\frac{u}{24} \bar{\phi}^4 \left(1 + \frac{3u}{32\pi^2} f^{(1, 0)}(rR^2, \mu R) + O(u^2)\right) + \\
    &\frac{u^3}{48}\bar{\phi}^6\left(\frac{R^2}{32\pi^2}f^{(2, 0)}(rR^2, \mu R) + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
\end{split}
\end{equation}
from which (\ref{eq:log_pdf}) is obtained by setting $g = \frac{3 u}{16
\pi^2}$ and $\mu^2 = r + R^{-2}$. This last choice is motivated as follows. The
subtraction in (\ref{eq:f_definition}) is similar to:
\begin{equation}
    \int_0^{\infty} \dd z\, \frac{e^{-a z} - e^{-b z}}{z} = \log \frac{b}{a}\,;
\end{equation}
when the asymptotic behavior of the two terms for large $z$ is not well
matched, the integral becomes large in magnitude, making the perturbative
expansion less reliable. With the choice $\mu^2 = r + R^{-2}$, both terms in
(\ref{eq:f_definition}) have the same asymptotic behavior $\sim e^{-z(\bar{r} +
1)}$, thus avoiding the large log problem over the widest possible range of
parameters.

@@

Curved manifolds allow for an additional renormalizable coupling: 
\begin{equation}
    \Delta S = \int_{\mathcal{M}} \frac{1}{2}\xi_0 \phi^2 \mathcal{R}\,,
\end{equation}
where $\mathcal{R}$ is the scalar curvature and $\xi_0$ is a dimensionless bare
coupling.  The free theory is Weyl invariant if $\xi_0 = \frac{1}{6}$. In the
presence of interactions, the coupling needs to be renormalized, and its
renormalized counterpart $\xi$ becomes a running coupling. At one loop, the
renormalization group equation for $\xi$ is:
\begin{equation}
    \mu \frac{\dd \xi}{\dd \mu} =  \frac{1}{3} g \left(\xi - \frac{1}{6}\right)\,.
\end{equation}

From this expression it seems that $\xi$ can be set to the critical value
$\frac{1}{6}$ at all energy scales. However, this turns out to be an illusion:
at higher orders in perturbation theory the renormalization group equation
becomes inhomogeneous \cite{Brown1980}. Therefore, on a curved manifold, $\xi$
is simply another free parameter of the scalar field theory.
